\subsection{Selective and Post-selection Inference}
\datum{Lecture 10: 02/05}
If assumptions are not correct, inference may not be valid.

\textbf{Linear model assumes:}
\begin{itemize}
    \item Effect of every $X_j$ is linear.
    \item $X_j$'s effects combine additively.
    \item All relevant covariates have been included.
    \item Noise is Gaussian (or at least not heavy-tailed).
    \item Noise terms have constant variance (regardless of values of $X$).
\end{itemize}

If we use data to help find setting in which linear model is true:
\[ 
\left.
\begin{array}{l}
\text{- Transformations of } X_j \\
\text{- Transformation of } Y \\
\text{- Selecting covariates to add/remove}
\end{array}
\right\} \to \begin{array}{c} \text{Problem of} \\ \text{selective inference /} \\ \text{post-selection} \\ \text{inference.} \end{array}
\]

\textbf{Note:} typically, ``selective'' / ``post-selection'' inference refers to choosing which covariates / hypotheses to test.

\textbf{Additional challenges to high-dimensional setting: ($n < p$)}
\begin{itemize}
    \item No assumptions on $\beta$ -- problem of identifiability.
    \item If we assume $\beta$ is sparse.
\end{itemize}

Typically, we need to assume sparsity level $k$ that is $\lesssim \frac{n}{\log p}$.
\begin{itemize}
    \item[$\hookrightarrow$] Can we estimate $\beta$ and perform inference on $\beta_j$'s, if we assume sparsity?
\end{itemize}

\textbf{Intuitively:} Select $k$ covariates to include, then perform inference on these $k$ $\beta_j$'s.
\begin{itemize}
    \item[$\hookrightarrow$] \textbf{Challenges:} (1) selection bias / selective inference
    \item[] \hspace{5.5mm} (2) interpretation given selected set
\end{itemize}

\begin{itemize}
    \item[$\hookrightarrow$] If we remove covariates from model, $\beta_j$ now represents association of $Y$ with $X_j$ when controlling for a \underline{reduced} set of other covariates.
\end{itemize}

\textbf{Concrete question: how to perform inference on $\beta_j$ using estimated $\hat{\beta}$ from Lasso}

\[ \hat{\beta} = \text{argmin} \left( \frac{1}{2} ||Y - X\beta||_2^2 + \lambda ||\beta||_1 \right) \]

Can we use $\hat{\beta}_j$ to test $\beta_j = 0$ or to build CI for $\beta_j$?

\textbf{Challenges:}
\begin{itemize}
    \item Selective inference choosing set of non-zero $\hat{\beta}_j$
    \item If $\beta_j \neq 0$, nonetheless may have $\hat{\beta}_j = 0$ or may underestimate magnitude $\mathbb{E}[\hat{\beta}_j] \neq \beta_j$ due to \textbf{shrinkage bias}
    \item Bias $\mathbb{E}[\hat{\beta}_j - \beta_j]$ hard to quantify: noise $\hat{\beta}_j - \mathbb{E}[\hat{\beta}_j]$ has unknown distribution
\end{itemize}

\textbf{An attempted solution:}
\begin{itemize}
    \item Run Lasso to select $S \subseteq [p]$ subset of covariates
    \item Then run least squares on reduced model (\textbf{severe selection bias})
\end{itemize}

\textbf{Data splitting --}
\begin{enumerate}
    \item Run Lasso on half the data to choose $S \subseteq [p]$
    \item Run last squares using other half of data.
\end{enumerate}

\textbf{Challenges:}
\begin{itemize}
    \item Reduced sample size
    \item By reducing the model, may lose validity of model or set of $X_j$, changing the interpretation of $\beta_j$
\end{itemize}

\textbf{If we do not split data:} Can control for multiplicity:
\begin{itemize}
    \item Build CI's via FCR controlling procedure
\end{itemize}

\emph{\textbf{Selective Inference}}

Lasso for choosing $S \subseteq [p]$ | Least squares on same data

For fixed $j \in [p]$ in linear regression:
\[ \hat{\beta}_j \sim \mathcal{N}\left( \beta_j, \sigma^2(X^TX)^{-1}_{jj} \right) \quad \text{e.g. under null: } \beta_j = 0: \]
\[ \hat{\beta}_j \sim \mathcal{N}\left( 0, \sigma^2(X^TX)^{-1}_{jj} \right) \xrightarrow{\text{calculate}} \mathbb{P}_{H_0}\{ |\hat{\beta}_j| \ge z^* \} \]

But, if $j \in [p]$ is chosen in a data dependent way
\[ \text{may be bigger } \hookrightarrow \mathbb{P}( |\hat{\beta}_j| \ge z^* \mid j \text{ is selected} ) \]

In general for any test statistic $T$,
\[ \text{Null distribution} \neq \text{Null distribution of } T \mid \text{selected event} \]
\[ \text{most severe if selection outcome is highly random / each possible outcome has low prob.} \]

\datum{Lecture 11: 02/10}
\textbf{Key challenge:} Distribution of $T$ under $H_0$ vs. Distribution of $T$ under $H_0$ conditional on the event that we choose to test $H_0$ with this $T$.
This may be $\neq$ in general and may be different if selection event is unlikely.

\textbf{General principle to correct for selection bias, our options include:}
\begin{itemize}
    \item Use data for inference that is $\perp\!\!\perp$ of selection (e.g. data splitting) or at least bound the independence.
    \item Use multiplicity correction (e.g. Bonferroni, FCR) to account for all possible choices that could have been selected.
    \item Perform inference using distribution conditional on the selection event
\end{itemize}

\begin{example}
    Applying selective inference strategy to linear regression.

    Distribution of data (before selection): $Y\sim\mathcal{N}(\mu,\sigma^2\text{I}_{n})$ wtih $\mu=X\beta$.

    We will consider selection events that can be described via an intersection of linear inequalities, i.e. events of the form $Y\in \{y\in\mathbb{R}^n\colon Ay\leq b\}$. We will test parameters of the form $\eta^\top \mu$ using test statistic $\eta^\top y$ for finitely many possible vectors $\eta \in \mathbb{R}^n$.

    Selection event of the form $Y\in \{y\in\mathbb{R}^n\colon Ay\leq b\}$ is sufficient to express selection procedures such as 
    \begin{itemize}
        \item Forward stepwise
        \item Marginal screening
        \item Lasso
        \item And others
    \end{itemize}
    \emph{\textbf{Note.}} In order to express the selection event in this form, may need to include additional information (e.g. order of selected covariates in forward stepwise).

    Selection procedure creates a partition (ignoring sets of measure 0) where $\mathbb{R}^n = \mathcal{Y}_1\cup \cdots \cup \mathcal{Y}_K$ where $\mathcal{Y}_k = \{y \colon A_ky\leq b_k\}$ with $A_k\in\mathbb{R}^{m_k\times n}$ and $b_k \in \mathbb{R}^{m_k}$.

    Distribution of $Y$ is $\sum_{k=1}^K\mathbb{P}[Y\in\mathcal{Y}_k] \cdot (\textup{Distribution of }Y\mid Y\in\mathcal{Y}_k)$. (Two-tailed test can be seen as selection inference conditioning on sign of $Y$)

    \textbf{Selective inference idea:} If $Y\in \mathcal{Y}_k$ determines the selection, then we can perform using $(\textup{Distribution of }Y\mid Y\in\mathcal{Y}_k)$.

    \textbf{What parameters to test?} Test parameters $\eta_{k,1}^\top\mu, \dots, \eta_{k,r_k}^\top\mu$, e.g. $\mathcal{Y}_k$ corresponds to choosing coordinates $X_1$ and $X_7$ and want to test coefficients on $X_1,X_7$ in the regression.

    In a low-dimensional setting $(p<n)$ least-squares gives $\hat{\beta} = (X^\top X)^{-1}X^\top Y$ and $\hat{\beta}_1 = [\underbrace{e_1 (X^\top X)^{-1}X^\top}_{\eta^\top}]Y$. Then $\eta^\top Y$ estimates $\eta^\top \mu$.

    In general (may have $p > n$). Partial regression coefficients: coefficient on $X_j$ in a submodel regressing $Y$ on a subset $S\subseteq [p]$ of covariates. If $S=\{1,7\}$, we want to find an $\eta$ that extracts the coefficients on $X_1$ in this regression.

    $$\hat{\beta}^S = (X_S^\top X_S)^{-1}X_S^\top Y$$ with $X_S \in \mathbb{R}^{n\times |S|}$ is a submatrix of $X$. And $\hat{\beta}_j^S$ is coefficient of $X_j$ in regression for submodel $S$ and is $\eta^\top Y$ where $\eta^\top = e_{j'}(X^\top_S X_S)^{-1} X^\top_s$ (and $j'$ is the new position of parameter $j$ in subset $S$).

    Interpreting these parameters: $\hat{\beta}_j^S$ estimates $$\beta^S_j = e_{j'}(X_S^\top X_S)^{-1}X_S^\top \mu = \beta_j + e_{j'}^\top (X_S^\top X_S)^{-1}X_S^\top (X_S\beta_S+X_{S^\textup{c}}\beta_{S^\textup{c}})=  \beta_j + e_{j'}^\top (X_S^\top X_S)^{-1} X_S^\top X_{S^c}\beta_{S^c}$$

    In linear regression setting:
    \begin{itemize}
        \item We want to select a set of partial regression coefficients to test $\beta_j^S$, i.e. select a model $S\subseteq [p]$ and test $\beta_j^S$ for each $j\in S$
        \item Selection event is of the form $\{y \colon Ay\leq b\}$
    \end{itemize}
    And we need to compute distribution of $\eta^\top  Y\mid AY\leq b$, i.e. distribution of $\hat{\beta}_j^S \mid \textup{select }S$
    \emph{\textbf{Note.}} Marginal distribution of $\eta^\top Y$ is $\mathcal{N}(\eta^\top \mu, \sigma^2\lVert\eta\rVert_2^2)$.
\end{example}


\datum{Lecture 12: 02/12}

\emph{\textbf{Selective Inference for Gaussian Linear Regression}}
\begin{itemize} 
    \item[(1)] Data $\sim$ distribution 
    \item[(2)] Select a target for inference + test statistic $T$
    \item[(3)] Compute distribution of $T$ conditional on (2)
\end{itemize}

For linear regression $Y\sim \mathcal{N}(\mu,\sigma^2 \textup{I}_n)$, the target is $\beta_j^S = e^\top_{j'}(X_S^\top X_S)^{-1}X_S^\top \mu = \eta^\top\mu$. The test statistic is $\hat{\beta}_j^S = \eta^\top Y$. What is the distribution of $\eta^\top Y \mid \textup{selected event}?$ We will consider selection events of the form $\{y\colon Ay\leq b \}$.

\textbf{Note.} If the selection event returns multiple inference questions, we would need to correct for multiplicity.

\textbf{Remaining Question.} What is the distribution of $\eta^{\top} Y \mid Ay \leq b$ without conditioning, i.e. $\eta^\top Y \sim \mathcal{N}(\underbrace{\eta^\top\mu}_{\text{target (unknown)}}, \underbrace{\sigma^2\lVert \eta\rVert_2^2}_{\text{known}})$. This does not depend on  $P_\eta^\perp(\mu)$.


\textbf{Problems:} In general, the distribution of $\eta^{\top} Y \mid Ay \leq b$ depends both on $\eta^\top \mu$ and $P^\perp_\eta (\mu)$, which is a $n-1$-dimensional nuisance parameter (\textit{Statistical}) Also, distribution has no simple expression and is hard to sample from (\textit{Computational}).

\textbf{Solution:} Condition on additional information $P_\eta^\perp(Y)$.

Data $Y\sim \mathcal{N}(\mu,\sigma^2\textup{I}_n)$. Identify which selection event contains $Y$, i.e. prespecified pattern $\mathbb{R}^n = \mathcal{Y}_1\cup\dots\cup\mathcal{Y}_K$ and $\mathcal{Y}_k = \{y\colon A_k y \leq b_k \}$ and identify the index $k$ such that $Y\in\mathcal{Y}_k$. $\mathcal{Y}_k$ determines the inference targets $\eta_{k,1}^\top\mu, \dots, \eta_{k,r_k}^\top \mu$. 

For each $j$, inference on $\eta^\top_{k,j}\mu$: condition on additional information $P^\perp_{\eta_{k,j}}(Y)$ and compute distribution of $\eta^\top_{k,j}Y \mid \{A_kY\leq b_K \}, P^\perp_{\eta_{k,j}}(Y)$ + use for inference. (\textbf{Note}: may need multiplicity correction)

\begin{example}
    Suppose we want to perform inference on $\mu_j$ where $j= \operatorname{argmax}_i|Y_i|$ (selection event) with $2n$ selection events $j^+, j^-$ for each $j=1,\dots,n$. For $j^+ \implies Y_j \geq 0$ and $|Y_j| \geq |Y_i| \quad \forall i\neq j \iff (Y_j \geq 0), Y_j\geq Y_i 
\wedge Y_j\geq -Y_i\quad \forall i\neq j$ ($2(n-1)$ equalities).
    
    For $j^+ \implies \eta=e_j, \quad \eta^\top\mu = \mu_j$.\\
    For $j^- \implies \eta=-e_j, \quad \eta^\top\mu = -\mu_j$.

For $j^+$, we have $P_y^\perp(Y) = (Y_1,\dots, Y_{j-1}, 0 , Y_{j+1},\dots, Y_n)$, the distribution of $Y_j \mid$ selection event $j^+$, is marginally $Y_j \sim \mathcal{N}(\mu_j, \sigma^2)$ and the conditional distribution of $Y_j$ is $TN(\mu_j, \sigma^2; \max_{i\neq j}|Y_i|. \infty)$ and for testing $H_{0,j}\colon \mu_j = 0$, we have p-value $\frac{1-\Phi\left(\frac{Y_j}{\sigma}\right)}{1-\Phi\left(\frac{\max_{i\neq j}|Y_i|}{\sigma}\right)}$ 
    
\end{example}

What is the conditional distribution of $\eta^\top Y \mid \{Ay\leq B\}, P_\eta^\perp(Y)$. Before conditioning $\eta Y \sim \mathcal{N}(\eta^\top Y, \sigma^2 \lVert\eta\rVert_2^2)$ and $\eta^\top Y \perp P_\eta^\perp(Y)$. We have
\begin{align*}
    Y = P_\eta(Y) + P_\eta^\perp(Y) = \eta^\top y \frac{\eta}{\lVert \eta\rVert_2^2}  + P_\eta^\perp(Y),
\end{align*}
which both are observed.
\begin{align*}
Ay \le b &\iff a_j^T y \le b_j \quad \forall j=1\dots m \\
&\iff a_j^T \left( \eta^T y \cdot \frac{\eta}{\|\eta\|_2^2} + P_\eta^\perp(y) \right) \le b_j \quad \forall j \\
&\iff \underset{\text{observed}}{\eta^T y} \cdot \frac{a_j^T \eta}{\|\eta\|_2^2} \le b_j - a_j^T P_\eta^\perp(y) \quad \forall j
\end{align*}

\begin{align*}
\text{i.e.} \quad &\forall j, \text{ with } a_j^T \eta > 0, \quad \eta^T y \le \frac{b_j - a_j^T P_\eta^\perp(y)}{a_j^T \eta / \|\eta\|_2^2} \\
&\forall j \text{ with } a_j^T \eta < 0, \quad \eta^T y \ge \dots \leftarrow \text{same}
\end{align*}

\[
\text{Let } R = \min_{j : a_j^T \eta > 0} \left\{ \frac{b_j - a_j^T P_\eta^\perp(y)}{a_j^T \eta / \|\eta\|_2^2} \right\} \text{ or } \infty \text{ if no such } j
\]

$\text{What is the conditional distribution of } \eta^T y \mid \{Ay \le b\}, P_\eta^\perp(y)?$

It is $TN(\eta^T \mu, \sigma^2 \|\eta\|_2^2; L, R) \quad \text{depend on } A, b, P_\eta^\perp(y) \text{ with p-value} = \frac{\Phi\left(\frac{R}{\sigma\|\eta\|_2}\right) - \Phi\left(\frac{\eta^T y}{\sigma\|\eta\|_2}\right)}{\Phi\left(\frac{R}{\sigma\|\eta\|_2}\right) - \Phi\left(\frac{L}{\sigma\|\eta\|_2}\right)}$

How to convert the Lasso selection procedure into an event of the form $\{y\colon Ay\leq b\}$?
\begin{itemize}
    \item fixed $\lambda > 0$
    \item $\text{sign}(\hat{\beta}_j)_{\text{lasso}}$ determines the selection, i.e. \begin{equation}
        \text{sign}(\hat{\beta}_j)_{\text{lasso}} =
    \begin{cases}
        +1 & X_j \text{ selected with +,}\\
        -1 & X_j \text{ selected with -,}\\
        0 & X_j \text{ not selected }.
    \end{cases}
    \end{equation} which we want to express in the form $\{Ay\leq B\}$.
\end{itemize}


